No meaningful effects of COVID-19 related social media use on well-being

### 2021-12-13 analyses.RmD
- added three studies to lit review

---
title: "Analyses"
output:
  html_document:
    toc: TRUE
    toc_float:
      collapsed: FALSE
    toc_depth: 2
    # code_folding: hide
bibliography      : 
  - "bibliography/r-references.bib"
  - "bibliography/references.bib"
  # pdf_document:
  #   toc: TRUE
---

```{r knitr-setup, include=F}
set.seed(170819)
knitr::opts_chunk$set(warning=F, echo=T, error=F, message=F, cache = T)
options(digits = 3, width = 200, scipen=999)
```

@@

```{r}
d_wide <- d_raw %>% 
  select(
    id = RESPID,
    gender = SD_GENDER,
    acc_bal = SD_ACCESS_BALCONY,
    acc_gar = SD_ACCESS_GARDEN,
    year_birth = SD_BIRTHYEAR,
    born_aus = SD_BORN_AUSTRIA,
    born_aus_prnts = SD_MIGR_BACKGR,
    county = SD_BULA,
    edu = SD_EDU,
    employment = SD_EMPLSTATUS_FEB2020,
    hh_adults = SD_HH_ADULTS,
    hh_child18 = SD_HH_CHILD18,
    hh_child17 = SD_HH_TEENS,
    hh_child14 = SD_HH_CHILD14,
    hh_child5 = SD_HH_CHILD5,
    hh_child2 = SD_HH_CHILD2,
    hh_oldfam = SD_HH_OLDERFAM,
    hh_outfam = SD_HH_OUTERFAM,
    hh_partner = SD_HH_PARTNER,
    # hh_income = SD_HHINCOME_FEB2020,
    home_sqm = SD_HOME_SQM,
    home_owner = SD_HOMEOWNER,
    # work_h = SD_WORKHOURS_FEB2020,
    health = all_of(health),
    life_sat = all_of(life_sat),
    aff_neg_1 = all_of(aff_neg_1),
    aff_neg_2 = all_of(aff_neg_2),
    aff_neg_3 = all_of(aff_neg_3),
    aff_neg_4 = all_of(aff_neg_4),
    aff_neg_5 = all_of(aff_neg_5),
    aff_neg_6 = all_of(aff_neg_6),
    aff_pos_1 = all_of(aff_pos_1),
    aff_pos_2 = all_of(aff_pos_2),
    aff_pos_3 = all_of(aff_pos_3),
    act_wrk = all_of(act_wrk),
    act_spo = all_of(act_spo),
    act_frn = all_of(act_frn),
    act_sho = all_of(act_sho),
    act_pet = all_of(act_pet),
    sat_dem = all_of(sat_dem),
    sat_dem = all_of(sat_dem),
    work_h = all_of(work_h),
    work_homeoff = all_of(work_homeoff),
    hh_income = all_of(hh_income),
    med_txt_kro = all_of(med_txt_kro),
    med_txt_sta = all_of(med_txt_sta),
    med_txt_pre = all_of(med_txt_pre),
    med_txt_oes = all_of(med_txt_oes),
    med_txt_kur = all_of(med_txt_kur),
    med_txt_slz = all_of(med_txt_slz),
    med_txt_son = all_of(med_txt_son),
    med_vid_orf = all_of(med_vid_orf),
    med_vid_pri = all_of(med_vid_pri),
    soc_med_fb = all_of(soc_med_fb),
    soc_med_tw = all_of(soc_med_tw),
    soc_med_ig = all_of(soc_med_ig),
    soc_med_yt = all_of(soc_med_yt),
    soc_med_wa = all_of(soc_med_wa),
    soc_med_like_share = all_of(soc_med_like_share),
    soc_med_read = all_of(soc_med_read),
    soc_med_post = all_of(soc_med_post),
    risk_prop = all_of(risk_prop),
    loc_cntrl_int_1 = all_of(loc_cntrl_int_1),
    loc_cntrl_int_2 = all_of(loc_cntrl_int_2),
    loc_cntrl_int_3 = all_of(loc_cntrl_int_3),
    loc_cntrl_int_4 = all_of(loc_cntrl_int_4)
  )

# rename variables
# names(d_wide) <- sub("\\.\\.\\.", "§", names(d_wide)) # replace "..." with "_"
```

@@

```{r}
d_wide %<>%
  mutate_at(vars(everything(.)), funs(na_if(., 88))) %>%
  mutate_at(vars(everything(.)), funs(na_if(., 99))) %>%
  mutate(
    male = 2 - .$gender,
    age = 2021 - .$year_birth,
    res_vienna = recode(.$county, `8` = 1L, .default = 0L,),
    born_aus = 2 - .$born_aus,
    home_owner = 2 - .$home_owner,
    employment_fac = factor(.$employment, 
                           # labels = c("Unselbststaendig erwerbstaetig (auch Lehrling), Privatwirtschaft",
                           #            "Unselbststaendig erwerbstaetig (auch Lehrling), Oeffentlicher Dienst",
                           #            "Selbststaendig erwerbstaetig",
                           #            "Arbeitslos",
                           #            "Pensionist*in",
                           #            "Ausschliesslich haushaltsfuehrend",
                           #            "Schueler/in, Student/in, Bildungskarenz",
                           #            "Sonstiges (Dauerhaft arbeitsunfaehig oder Praesenz-/Zivildiener)",
                           #            "In Elternkarenz")
                           labels = c("Unemployed",
                                      "Industrie",
                                      "Public service",
                                      "Self-employed",
                                      "Retired",
                                      "Housekeeping",
                                      "Student",
                                      "Incapacitated",
                                      "Parental Leave"),
                            levels = c(4, 1:3, 5:8, 10) # make unemployment reference cat
                           ),
    edu_fac = factor(.$edu,
                     # labels = c("Volksschule oder weniger", 
                     #            "Hauptschule oder AHS Unterstufe",
                     #            "Polytechnikum, BMS (Fachschule, z.B. HASCH)",
                     #            "Lehre, Berufsschule",
                     #            "AHS mit Matura",
                     #            "BHS mit Matura (z.B. HTL, HAK, HBLA etc.)",
                     #            "Hochschulverwandte Lehranstalt oder Kolleg",
                     #            "Bachelor",
                     #            "Magister / Master / Diplomingenieur / Fachhochschule",
                     #            "Doktor / PhD"),
                     labels = c("No degree", 
                                "Middle school",
                                "Vocational school",
                                "Technical school",
                                "High school",
                                "Applied high school",
                                "State college",
                                "Bachelor",
                                "Master",
                                "PhD")
                     ))
```

@@

<!-- Filter waves without data -->

<!-- ```{r eval=F} -->
<!-- d_long_test <- d_long %>% -->
<!--   filter(!is.na(.[vars_varying])) -->
<!-- ``` -->

Recode values

@@

<!-- Create vectors with names -->

<!-- ```{r} -->
<!-- # create vector of names -->
<!-- vars_varying <- names(select(d_long, "health_b":"loc_cntrl_int_4_w")) -->
<!-- predictors_varying <- select(d_long, "health_b":"loc_cntrl_int_4_w") %>%  -->
<!--   select(-starts_with("aff_"), -starts_with("life_")) %>%  -->
<!--   names() -->
<!-- ``` -->

@@

The data fit the model very well, `#td::fit_txt(cfa_aff_pos)`.
Let's next inspect reliability

@@

The data fit the model very well, `#td::fit_txt(cfa_aff_neg)`.

@@

```{r}
model_soc_med_tw <- lmer(soc_med_tw ~ (1 | id) + (1 | wave), d_long_imp)
summary(model_soc_med_tw)

# Bayesian Approach
# model_soc_med_tw <- brm(soc_med_tw ~ (1 | id) + (1 | wave), d_long_imp
# Also throws a warning
# model_soc_med_tw <- brm(soc_med_tw ~ (1 | id) + (1 | wave), d_long_imp, control=list(adapt_delta=0.82))
# Currently doesn't work, likely for lack of memory
```

@@

The data fit the model very well, `#td::fit_txt(cfa_loc_cntrl_int)`.

@@

```{r}
model_aff_pos_lmer <- lmerTest::lmer(aff_pos_fs ~ 
                                        (1 | id) + (1 | wave) + 
                                        soc_med_read_w + soc_med_like_share_w + soc_med_post_w + 
                                        soc_med_fb_w + soc_med_ig_w + soc_med_wa_w + soc_med_yt_w + soc_med_tw_w +
                                        soc_med_read_b + soc_med_like_share_b + soc_med_post_b + 
                                        soc_med_fb_b + soc_med_ig_b + soc_med_wa_b + soc_med_yt_b + soc_med_tw_b +  
                                        age + male + born_aus + born_aus_prnts + edu_fac + employment_fac + health_b + 
                                        res_vienna + acc_bal + acc_gar + home_sqm + 
                                        # hh_income_b + # large dropout (900 obs)
                                        # hh_adults + # too much dropout
                                        # hh_child18 + hh_child17 + hh_child14 + hh_child5 + hh_child2 + hh_oldfam + hh_outfam + hh_partner + 
                                        # home_owner + # too much dropout
                                        med_txt_kro_b + med_txt_sta_b + med_txt_pre_b + med_txt_oes_b + med_txt_kur_b + med_txt_slz_b + med_txt_son_b + 
                                        med_vid_orf_b + med_vid_pri_b + 
                                        risk_prop_b + loc_cntrl_int_m_b + 
                                        act_wrk_w + act_spo_w + act_frn_w + act_sho_w + act_pet_w + 
                                        act_wrk_b + act_spo_b + act_frn_b + act_sho_b + act_pet_b +
                                        sat_dem_w + sat_dem_b 
                                      , data = d_long_imp)
summary(model_aff_pos_lmer)
```

@@

```{r}
model_aff_neg_lmer <- lmerTest::lmer(aff_neg_fs ~ 
                                        (1 | id) + (1 | wave) + 
                                        soc_med_read_w + soc_med_like_share_w + soc_med_post_w + 
                                        soc_med_fb_w + soc_med_ig_w + soc_med_wa_w + soc_med_yt_w + soc_med_tw_w +
                                        soc_med_read_b + soc_med_like_share_b + soc_med_post_b + 
                                        soc_med_fb_b + soc_med_ig_b + soc_med_wa_b + soc_med_yt_b + soc_med_tw_b +  
                                        age + male + born_aus + born_aus_prnts + edu_fac + employment_fac + health_b + 
                                        res_vienna + acc_bal + acc_gar + home_sqm + 
                                        # hh_income_b + # large dropout (900 obs)
                                        # hh_adults + # too much dropout
                                        # hh_child18 + hh_child17 + hh_child14 + hh_child5 + hh_child2 + hh_oldfam + hh_outfam + hh_partner + 
                                        # home_owner + # too much dropout
                                        med_txt_kro_b + med_txt_sta_b + med_txt_pre_b + med_txt_oes_b + med_txt_kur_b + med_txt_slz_b + med_txt_son_b + 
                                        med_vid_orf_b + med_vid_pri_b + 
                                        risk_prop_b + loc_cntrl_int_m_b + 
                                        act_wrk_w + act_spo_w + act_frn_w + act_sho_w + act_pet_w + 
                                        act_wrk_b + act_spo_b + act_frn_b + act_sho_b + act_pet_b +
                                        sat_dem_w + sat_dem_b 
                                      , data = d_long_imp)
summary(model_aff_neg_lmer)
```

>>>>

---
title: "Analyses"
output:
  html_document:
    toc: TRUE
    toc_float:
      collapsed: FALSE
    toc_depth: 2
    code_folding: hide
bibliography      : 
  - "bibliography/r-references.bib"
  - "bibliography/references.bib"
  # pdf_document:
  #   toc: TRUE
---

```{r knitr-setup, include=F}
set.seed(170819)
knitr::opts_chunk$set(warning=F, echo=T, error=F, message=F, cache = T)
options(digits = 3, width = 250, scipen=999)
```

Here you can find the code and the results of all analyses. To see the code, click on button "Code". Alternatively, you can download the rmd file from the github repo.

@@

```{r}
d_wide <- d_raw %>% 
  select(
    id = RESPID,
    gender = SD_GENDER,
    acc_bal = SD_ACCESS_BALCONY,
    acc_gar = SD_ACCESS_GARDEN,
    year_birth = SD_BIRTHYEAR,
    born_aus = SD_BORN_AUSTRIA,
    born_aus_prnts = SD_MIGR_BACKGR,
    county = SD_BULA,
    edu = SD_EDU,
    employment = SD_EMPLSTATUS_FEB2020,
    hh_adults = SD_HH_ADULTS,
    hh_child18 = SD_HH_CHILD18,
    hh_child17 = SD_HH_TEENS,
    hh_child14 = SD_HH_CHILD14,
    hh_child5 = SD_HH_CHILD5,
    hh_child2 = SD_HH_CHILD2,
    hh_oldfam = SD_HH_OLDERFAM,
    hh_outfam = SD_HH_OUTERFAM,
    hh_partner = SD_HH_PARTNER,
    # hh_income = SD_HHINCOME_FEB2020,
    home_sqm = SD_HOME_SQM,
    home_owner = SD_HOMEOWNER,
    # work_h = SD_WORKHOURS_FEB2020,
    health = all_of(health),
    life_sat = all_of(life_sat),
    aff_neg_1 = all_of(aff_neg_1),
    aff_neg_2 = all_of(aff_neg_2),
    aff_neg_3 = all_of(aff_neg_3),
    aff_neg_4 = all_of(aff_neg_4),
    aff_neg_5 = all_of(aff_neg_5),
    aff_neg_6 = all_of(aff_neg_6),
    aff_pos_1 = all_of(aff_pos_1),
    aff_pos_2 = all_of(aff_pos_2),
    aff_pos_3 = all_of(aff_pos_3),
    act_wrk = all_of(act_wrk),
    act_spo = all_of(act_spo),
    act_frn = all_of(act_frn),
    act_sho = all_of(act_sho),
    act_pet = all_of(act_pet),
    sat_dem = all_of(sat_dem),
    sat_dem = all_of(sat_dem),
    work_h = all_of(work_h),
    work_homeoff = all_of(work_homeoff),
    hh_income = all_of(hh_income),
    med_txt_kro = all_of(med_txt_kro),
    med_txt_sta = all_of(med_txt_sta),
    med_txt_pre = all_of(med_txt_pre),
    med_txt_oes = all_of(med_txt_oes),
    med_txt_kur = all_of(med_txt_kur),
    med_txt_slz = all_of(med_txt_slz),
    med_txt_son = all_of(med_txt_son),
    med_vid_orf = all_of(med_vid_orf),
    med_vid_pri = all_of(med_vid_pri),
    soc_med_fb = all_of(soc_med_fb),
    soc_med_tw = all_of(soc_med_tw),
    soc_med_ig = all_of(soc_med_ig),
    soc_med_yt = all_of(soc_med_yt),
    soc_med_wa = all_of(soc_med_wa),
    soc_med_like_share = all_of(soc_med_like_share),
    soc_med_read = all_of(soc_med_read),
    soc_med_post = all_of(soc_med_post),
    risk_prop = all_of(risk_prop),
    loc_cntrl_int_1 = all_of(loc_cntrl_int_1),
    loc_cntrl_int_2 = all_of(loc_cntrl_int_2),
    loc_cntrl_int_3 = all_of(loc_cntrl_int_3),
    loc_cntrl_int_4 = all_of(loc_cntrl_int_4)
  )
```

@@

```{r}
d_wide %<>%
  mutate_at(vars(everything(.)), funs(na_if(., 88))) %>%
  mutate_at(vars(everything(.)), funs(na_if(., 99))) %>%
  mutate(
    male = 2 - .$gender,
    age = 2021 - .$year_birth,
    res_vienna = recode(.$county, `8` = 1L, .default = 0L,),
    born_aus = 2 - .$born_aus,
    home_owner = 2 - .$home_owner,
    employment_fac = factor(.$employment, 
                           labels = c("Unemployed",
                                      "Industrie",
                                      "Public service",
                                      "Self-employed",
                                      "Retired",
                                      "Housekeeping",
                                      "Student",
                                      "Incapacitated",
                                      "Parental Leave"),
                            levels = c(4, 1:3, 5:8, 10) # make unemployment reference cat
                           ),
    edu_fac = factor(.$edu,
                     labels = c("No degree", 
                                "Middle school",
                                "Vocational school",
                                "Technical school",
                                "High school",
                                "Applied high school",
                                "State college",
                                "Bachelor",
                                "Master",
                                "PhD")
                     ))
```

@@

Recode values.

@@

@@

The data fit the model very well, `r td::fit_txt(cfa_aff_pos)`.
Let's next inspect reliability

@@

The data fit the model very well, `r td::fit_txt(cfa_aff_neg)`.

@@

```{r}
model_soc_med_tw <- lmer(soc_med_tw ~ (1 | id) + (1 | wave), d_long_imp)
summary(model_soc_med_tw)
```

@@

The data fit the model very well, `r td::fit_txt(cfa_loc_cntrl_int)`.

@@

```{r}
model_aff_pos_lmer <- lmerTest::lmer(aff_pos_fs ~ 
                                        (1 | id) + (1 | wave) + 
                                        soc_med_read_w + soc_med_like_share_w + soc_med_post_w + 
                                        soc_med_fb_w + soc_med_ig_w + soc_med_wa_w + soc_med_yt_w + soc_med_tw_w +
                                        soc_med_read_b + soc_med_like_share_b + soc_med_post_b + 
                                        soc_med_fb_b + soc_med_ig_b + soc_med_wa_b + soc_med_yt_b + soc_med_tw_b +  
                                        age + male + born_aus + born_aus_prnts + edu_fac + employment_fac + health_b + 
                                        res_vienna + acc_bal + acc_gar + home_sqm + 
                                        med_txt_kro_b + med_txt_sta_b + med_txt_pre_b + med_txt_oes_b + med_txt_kur_b + med_txt_slz_b + med_txt_son_b + 
                                        med_vid_orf_b + med_vid_pri_b + 
                                        risk_prop_b + loc_cntrl_int_m_b + 
                                        act_wrk_w + act_spo_w + act_frn_w + act_sho_w + act_pet_w + 
                                        act_wrk_b + act_spo_b + act_frn_b + act_sho_b + act_pet_b +
                                        sat_dem_w + sat_dem_b 
                                      , data = d_long_imp)
summary(model_aff_pos_lmer)
```

@@

```{r}
model_aff_neg_lmer <- lmerTest::lmer(aff_neg_fs ~ 
                                        (1 | id) + (1 | wave) + 
                                        soc_med_read_w + soc_med_like_share_w + soc_med_post_w + 
                                        soc_med_fb_w + soc_med_ig_w + soc_med_wa_w + soc_med_yt_w + soc_med_tw_w +
                                        soc_med_read_b + soc_med_like_share_b + soc_med_post_b + 
                                        soc_med_fb_b + soc_med_ig_b + soc_med_wa_b + soc_med_yt_b + soc_med_tw_b +  
                                        age + male + born_aus + born_aus_prnts + edu_fac + employment_fac + health_b + 
                                        res_vienna + acc_bal + acc_gar + home_sqm + 
                                        med_txt_kro_b + med_txt_sta_b + med_txt_pre_b + med_txt_oes_b + med_txt_kur_b + med_txt_slz_b + med_txt_son_b + 
                                        med_vid_orf_b + med_vid_pri_b + 
                                        risk_prop_b + loc_cntrl_int_m_b + 
                                        act_wrk_w + act_spo_w + act_frn_w + act_sho_w + act_pet_w + 
                                        act_wrk_b + act_spo_b + act_frn_b + act_sho_b + act_pet_b +
                                        sat_dem_w + sat_dem_b 
                                      , data = d_long_imp)
summary(model_aff_neg_lmer)
```

### 2021-12-13 analyses_additional.Rmd
- added three studies to lit review

---
title: "Additional analyses"
output:
  html_document:
    toc: TRUE
    toc_float:
      collapsed: FALSE
    toc_depth: 2
    # code_folding: hide
bibliography      : 
  - "bibliography/r-references.bib"
  - "bibliography/references.bib"
  # pdf_document:
  #   toc: TRUE
---

@@

@@

>>>>

---
title: "Additional analyses"
output:
  html_document:
    toc: TRUE
    toc_float:
      collapsed: FALSE
    toc_depth: 2
    code_folding: hide
bibliography      : 
  - "bibliography/r-references.bib"
  - "bibliography/references.bib"
  # pdf_document:
  #   toc: TRUE
---

@@

Here you can find the code and the results of all additional analyses. To see the code, click on button "Code". Alternatively, you can download the rmd file from the github repo.

@@

Save results so that we can extract them in the manuscript.

### 2021-12-13 manuscript.Rmd
- added three studies to lit review

@@

## Defining Well-being and Media Use

@@

So far, there is only little empirical research on COVID-19 related social media use on well-being.
In their study on the relations between media use and mental health during the pandemic, @bendauAssociationsCOVID19Related2021 found that people who used social media as a primary source of information reported on average "significantly more unspecific anxiety and depression [] and significantly more specific COVID-19 related anxiety symptoms" (p. 288).
@edenMediaCopingCOVID192020 analyzed the media use of 425 US college students during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, @choiMediatedCommunicationMatters2021 reported that people who used media to attain information were more lonely and less satisfied with their lives.
@stainbackCOVID1924News2020 analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
Together, the literature emphasizes potentially negative effects of social media as news use [see also @liuRelationOfficialWhatsAppdistributed2020; @riehmAssociationsMediaExposure2020].
However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below).
We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education.

@@

First, one could assume a _direct_ negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Repeatedly consuming such news can be depressing, perhaps even changing some general perspectives on life.
<!-- So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well. -->
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects.
<!-- However, in light of a worldwide pandemic with millions of deaths, the negative direct effect seems more plausible. -->

@@

On the one hand, the results are not aligned with several recent studies analyzing similar or closely related research questions.
This includes a study by @bendauAssociationsCOVID19Related2021, which showed negative relations between social media and well-being.
However, note that @bendauAssociationsCOVID19Related2021 analyzed cross-sectional data on a between-person level while not controlling for third variables, which is not optimal for investigating causal effects.
On the other hand, the results are well-aligned with recent studies and meta-analyses analyzing the effects of social media use from a more general perspective or from a somewhat different angle. 
These studies have found that the effects of various types of social media use on several well-being indicators are small at best, often too small to matter [@meierComputermediatedCommunicationSocial2020a; @orbenTeenagersScreensSocial2020; @fergusonThisMetaanalysisScreen2021], which echoes the results obtained here.

@@

>>>>

note: |
  Preprint submitted for publication. Please cite carefully.

@@

## Understanding Well-being and Media Use

@@

So far, there is only little empirical research on COVID-19 related social media use on well-being.
In their study on the relations between media use and mental health during the pandemic, @bendauAssociationsCOVID19Related2021 found that people who used social media as a primary source of information reported on average "significantly more unspecific anxiety and depression [] and significantly more specific COVID-19 related anxiety symptoms" (p. 288).
@edenMediaCopingCOVID192020 analyzed the media use of 425 US college students during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, @choiMediatedCommunicationMatters2021 reported that people who used media to attain information were more lonely and less satisfied with their lives.
@stainbackCOVID1924News2020 analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
A four-wave panel study with 384 young adults from the U.S. analyzed the effects of general digital technology use---objectively measured via screenshots of screen-time applications---on mental health, separating within- and between-person relations [@sewallObjectivelyMeasuredDigital2021].
The results showed that digital technology did not have any significant effects on mental health [for a similar study with virtually identical results, see @bradleyStressMoodSmartphone2021].
Together, the literature is mixed, with a slight focus on the negative effects of social media as news use [see also dornemannHowGoodBad2021; @liuRelationOfficialWhatsAppdistributed2020; @riehmAssociationsMediaExposure2020].
However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below).
We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education.

@@

First, one could assume a _direct_ negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Consuming such news can be depressing [@dornemannHowGoodBad2021], perhaps even changing some general perspectives on life.
<!-- So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well. -->
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects [@dornemannHowGoodBad2021].
<!-- However, in light of a worldwide pandemic with millions of deaths, the negative direct effect seems more plausible. -->

@@

On the one hand, the results are not aligned with several recent studies analyzing similar or closely related research questions.
This includes a study by @bendauAssociationsCOVID19Related2021, which showed negative relations between social media and well-being [but see @bradleyStressMoodSmartphone2021; or @sewallObjectivelyMeasuredDigital2021].
However, note that @bendauAssociationsCOVID19Related2021 analyzed cross-sectional data on a between-person level while not controlling for third variables, which is not optimal for investigating causal effects.
On the other hand, the results are well-aligned with recent studies and meta-analyses analyzing the effects of social media use from a more general perspective or from a somewhat different angle. 
These studies have found that the effects of various types of social media use on several well-being indicators are small at best, often too small to matter [@meierComputermediatedCommunicationSocial2020a; @orbenTeenagersScreensSocial2020; @fergusonThisMetaanalysisScreen2021], which echoes the results obtained here.

@@

# Acknowledgements

I would like to thank Niklas Johannes for providing valuable feedback on this manuscript.

### 2021-12-13 manuscript.tex
- added three studies to lit review

@@

\hypertarget{defining-well-being-and-media-use}{%
\subsection{Defining Well-being and Media Use}\label{defining-well-being-and-media-use}}

@@

So far, there is only little empirical research on COVID-19 related social media use on well-being.
In their study on the relations between media use and mental health during the pandemic, Bendau et al. (2021) found that people who used social media as a primary source of information reported on average ``significantly more unspecific anxiety and depression {[}{]} and significantly more specific COVID-19 related anxiety symptoms'' (p.~288).
Eden, Johnson, Reinecke, and Grady (2020) analyzed the media use of 425 US college students during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, Choi and Choung (2021) reported that people who used media to attain information were more lonely and less satisfied with their lives.
Stainback, Hearne, and Trieu (2020) analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
Together, the literature emphasizes potentially negative effects of social media as news use (see also Liu \& Tong, 2020; Riehm et al., 2020).
However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below).
We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education.

@@

First, one could assume a \emph{direct} negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Repeatedly consuming such news can be depressing, perhaps even changing some general perspectives on life.
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects.

@@

On the one hand, the results are not aligned with several recent studies analyzing similar or closely related research questions.
This includes a study by Bendau et al. (2021), which showed negative relations between social media and well-being.
However, note that Bendau et al. (2021) analyzed cross-sectional data on a between-person level while not controlling for third variables, which is not optimal for investigating causal effects.
On the other hand, the results are well-aligned with recent studies and meta-analyses analyzing the effects of social media use from a more general perspective or from a somewhat different angle.
These studies have found that the effects of various types of social media use on several well-being indicators are small at best, often too small to matter (Ferguson et al., 2021; Meier \& Reinecke, 2020; Orben, 2020), which echoes the results obtained here.

@@

@@

@@

@@

>>>>

\note{

Preprint submitted for publication. Please cite carefully.

@@

\hypertarget{understanding-well-being-and-media-use}{%
\subsection{Understanding Well-being and Media Use}\label{understanding-well-being-and-media-use}}

@@

So far, there is only little empirical research on COVID-19 related social media use on well-being.
In their study on the relations between media use and mental health during the pandemic, Bendau et al. (2021) found that people who used social media as a primary source of information reported on average ``significantly more unspecific anxiety and depression {[}{]} and significantly more specific COVID-19 related anxiety symptoms'' (p.~288).
Eden, Johnson, Reinecke, and Grady (2020) analyzed the media use of 425 US college students during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, Choi and Choung (2021) reported that people who used media to attain information were more lonely and less satisfied with their lives.
Stainback, Hearne, and Trieu (2020) analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
A four-wave panel study with 384 young adults from the U.S. analyzed the effects of general digital technology use---objectively measured via screenshots of screen-time applications---on mental health, separating within- and between-person relations (Sewall, Goldstein, \& Rosen, 2021).
The results showed that digital technology did not have any significant effects on mental health (for a similar study with virtually identical results, see Bradley \& Howard, 2021).
Together, the literature is mixed, with a slight focus on the negative effects of social media as news use {[}see also dornemannHowGoodBad2021; Liu and Tong (2020); Riehm et al. (2020){]}.
However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below).
We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education.

@@

First, one could assume a \emph{direct} negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Consuming such news can be depressing (Dörnemann, Boenisch, Schommer, Winkelhorst, \& Wingen, 2021), perhaps even changing some general perspectives on life.
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects (Dörnemann, Boenisch, Schommer, Winkelhorst, \& Wingen, 2021).

@@

On the one hand, the results are not aligned with several recent studies analyzing similar or closely related research questions.
This includes a study by Bendau et al. (2021), which showed negative relations between social media and well-being (but see Bradley \& Howard, 2021; or Sewall, Goldstein, \& Rosen, 2021).
However, note that Bendau et al. (2021) analyzed cross-sectional data on a between-person level while not controlling for third variables, which is not optimal for investigating causal effects.
On the other hand, the results are well-aligned with recent studies and meta-analyses analyzing the effects of social media use from a more general perspective or from a somewhat different angle.
These studies have found that the effects of various types of social media use on several well-being indicators are small at best, often too small to matter (Ferguson et al., 2021; Meier \& Reinecke, 2020; Orben, 2020), which echoes the results obtained here.

@@

\leavevmode\hypertarget{ref-bradleyStressMoodSmartphone2021}{}%
Bradley, A., \& Howard, A. (2021). \emph{Stress, mood, and smartphone use in {University} students: {A} 12-week longitudinal study}. OSF Preprints. \url{https://doi.org/10.31219/osf.io/frvpb}

@@

\leavevmode\hypertarget{ref-dornemannHowGoodBad2021}{}%
Dörnemann, A., Boenisch, N., Schommer, L., Winkelhorst, L., \& Wingen, T. (2021). \emph{How do good and bad news impact mood during the {Covid}-19 pandemic? {The} role of similarity}. OSF Preprints. \url{https://doi.org/10.31219/osf.io/sy2kd}

@@

\leavevmode\hypertarget{ref-sewallObjectivelyMeasuredDigital2021}{}%
Sewall, C. J. R., Goldstein, T. R., \& Rosen, D. (2021). Objectively measured digital technology use during the {COVID}-19 pandemic: {Impact} on depression, anxiety, and suicidal ideation among young adults. \emph{Journal of Affective Disorders}, \emph{288}, 145--147. \url{https://doi.org/10.1016/j.jad.2021.04.008}

@@

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

I would like to thank Niklas Johannes for providing valuable feedback on this manuscript.
